\documentclass[11pt]{article}

% ---------- Packages ----------
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage[most]{tcolorbox}
\usepackage{titlesec}
\usepackage{parskip}
\usepackage{hyperref}
\usepackage{fancyhdr}

% ---------- Hyperref ----------
\hypersetup{
  colorlinks=true,
  linkcolor=blue!50!black,
  urlcolor=blue!50!black,
  pdftitle={AP Statistics: AI Tutoring, Compliance, and Evidence (v2)}
}

% ---------- Header / footer ----------
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small AP Statistics --- AI Tutoring, Compliance, and Evidence (v2)}
\fancyhead[R]{\small \thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ---------- Section formatting ----------
\titleformat{\part}[display]
  {\normalfont\Huge\bfseries\centering}
  {\partname\ \thepart}{0.5em}{}
\titleformat{\section}
  {\normalfont\Large\bfseries\color{blue!50!black}}{\thesection}{1em}{}
\titleformat{\subsection}
  {\normalfont\large\bfseries\color{blue!40!black}}{\thesubsection}{1em}{}

% ---------- Boxes ----------
\newtcolorbox{scenario}{
  colback=gray!5, colframe=gray!50,
  title=Scenario, fonttitle=\bfseries,
  boxrule=0.5pt, arc=2pt
}
\newtcolorbox{notebox}{
  colback=blue!3, colframe=blue!40!black,
  title=Note, fonttitle=\bfseries,
  boxrule=0.5pt, arc=2pt
}
\newtcolorbox{expectedbox}{
  colback=green!3, colframe=green!40!black,
  title=Expected Response, fonttitle=\bfseries,
  boxrule=0.5pt, arc=2pt
}
\newtcolorbox{keybox}{
  colback=yellow!8, colframe=orange!60!black,
  title=Worked Answer Key, fonttitle=\bfseries,
  boxrule=0.5pt, arc=2pt
}
\newtcolorbox{cautionbox}{
  colback=red!3, colframe=red!50!black,
  title=Cautionary Evidence, fonttitle=\bfseries,
  boxrule=0.5pt, arc=2pt
}

% ---------- Custom commands ----------
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\phat}{\hat{p}}
\newcommand{\xbar}{\bar{x}}

\title{\bfseries AP Statistics:\\ AI Tutoring, Compliance, and Evidence \\[0.4em]
\large A Combined Capstone and Investigative Task \\[0.2em]
\normalsize Version 2 (corrected)}
\author{}
\date{}

\begin{document}
\maketitle
\thispagestyle{empty}

\begin{center}
\begin{minipage}{0.85\textwidth}
\itshape
This document combines two complementary AP Statistics investigations.
\textbf{Part I} is a 60-question synthetic capstone covering Units 1--9,
designed to assess every major inferential and conceptual skill in the
course. \textbf{Part II} is a focused investigative task that applies
those same tools to a real-world causal question about AI-tutor
pedagogy, drawing on a published 2025 randomized controlled trial.
Together they distinguish \emph{intent-to-treat effects},
\emph{on-treatment associations}, \emph{confounding}, and
\emph{selection bias}.
\end{minipage}
\end{center}

\vspace{1em}

\begin{notebox}
\textbf{Defensible thesis (refined).}
\emph{Scaffolded AI-as-tutor --- AI permitted during the learning
phase with mandatory independent demonstration on assessments
(handwritten reasoning, verbal defense, no-AI in-class checks) ---
produces faster skill acquisition than either prohibition or
unscaffolded answer-engine use.} The published evidence supports
this conditional claim, not the stronger claim that all AI access
dominates all alternatives.
\end{notebox}

\tableofcontents
\newpage

% ============================================================
\part{Capstone: AI Tutoring, Compliance, and Evidence}
% ============================================================

\begin{notebox}
This is a synthetic AP Statistics investigation. The scenario is
realistic, but the data are invented so the statistical reasoning can
be evaluated cleanly.
\end{notebox}

\section{Scenario}

A district is debating whether AI tutoring should be allowed during the
learning phase of AP Statistics. Some teachers argue that AI mainly
enables cheating. Others argue that AI works like a private tutor:
useful during practice, but not a substitute for independent reasoning
on assessments.

The district runs a six-week study with 240 AP Statistics students
from 8 schools. Within each school, students are randomly assigned to
one of two instructional conditions:

\begin{itemize}[leftmargin=*]
  \item \textbf{AI-access condition:} Students may use an AI tutor
  during practice, homework, and review. They must still show
  handwritten work and complete in-class checks without AI.
  \item \textbf{Standard-resource condition:} Students use AP
  Classroom videos, teacher notes, textbook examples, and ordinary
  peer discussion. They may not use AI during the study period.
\end{itemize}

Researchers measure each student's pretest score, posttest score,
daily AI-tutor minutes, phone-policy environment, engagement rating,
and whether the student passes a later AP-style practice exam. Use a
significance level of $\alpha = 0.05$ unless otherwise stated.

\section{Data}

\subsection*{Table A: One-Variable Data}
For 20 randomly selected students in the AI-access condition, the
recorded average daily AI-tutor minutes were:
\[
0,\ 0,\ 5,\ 8,\ 10,\ 12,\ 15,\ 18,\ 22,\ 25,\ 30,\ 32,\ 35,\ 40,\ 45,\ 50,\ 55,\ 60,\ 75,\ 95
\]

\subsection*{Table B: Randomized Study Summary}
\begin{center}
\small
\begin{tabular}{lrrrrrrrr}
\toprule
Condition & $n$ & Mean pre & SD pre & Mean post & SD post & Mean gain & SD gain & Passed \\
\midrule
AI-access          & 120 & 48.2 & 13.8 & 66.8 & 15.6 & 18.6 & 12.4 & 69 \\
Standard-resource  & 120 & 48.9 & 14.1 & 58.7 & 16.2 &  9.8 & 11.7 & 47 \\
\bottomrule
\end{tabular}
\end{center}

\subsection*{Table C: Actual AI Use Within the AI-Access Condition}
\begin{center}
\small
\begin{tabular}{lrrrr}
\toprule
Actual behavior & $n$ & Mean gain & SD gain & Passed \\
\midrule
Regular use ($\geq 20$ min/day)        & 76 & 23.4 & 11.6 & 55 \\
Little or no use ($< 20$ min/day)      & 44 & 10.3 & 10.9 & 14 \\
\bottomrule
\end{tabular}
\end{center}

\subsection*{Table D: Districtwide Observational Survey of AI Culture and Use}

\begin{notebox}
\textbf{Important:} Table D is a \emph{separate} observational survey
conducted across the district, not part of the randomized study. It
samples students broadly --- including those outside the six-week
trial --- to characterize the relationship between school culture and
self-reported AI use. It cannot be used to draw causal conclusions
about AI access.
\end{notebox}

\begin{center}
\small
\begin{tabular}{lrrrr}
\toprule
School culture & Regular & Occasional & None & Total \\
\midrule
Restrictive phone/AI culture & 21 & 39 & 60 & 120 \\
Open AI-tutor culture        & 66 & 34 & 20 & 120 \\
\midrule
\textbf{Total}               & \textbf{87} & \textbf{73} & \textbf{80} & \textbf{240} \\
\bottomrule
\end{tabular}
\end{center}

\subsection*{Table E: Regression Summary (AI-Access Subset)}
For 80 students in the \textbf{AI-access condition} with complete
usage logs, researchers fit a least-squares regression model
predicting gain score from average daily AI-tutor minutes:
\[
\widehat{\text{gain}} = 5.8 + 0.40(\text{minutes})
\]
Additional information:
\begin{itemize}[leftmargin=*,nosep]
  \item $n = 80$
  \item $r = 0.62$
  \item $s = 9.8$
  \item $SE_b = 0.061$ (standard error of the slope)
  \item Residual plot shows random scatter with roughly constant vertical spread.
  \item Histogram of residuals is roughly unimodal and approximately symmetric.
\end{itemize}

\subsection*{Table F: Probability Model}
Based on a larger district survey, suppose an AP Statistics student
independently has probability $p = 0.32$ of becoming a regular
AI-tutor user when AI access is allowed but not required.

\section{Questions}

\subsection*{Unit 1: Exploring One-Variable Data}
\begin{enumerate}[leftmargin=*]
\item Classify each variable as categorical or quantitative:
instructional condition, school culture, average daily AI-tutor
minutes, gain score, passed AP-style practice exam.
\item Use Table A to make an appropriate graph of AI-tutor minutes.
Describe shape, center, spread, and unusual features in context.
\item Find the median, $Q_1$, $Q_3$, IQR, and range of Table A.
\item Use the $1.5 \times \text{IQR}$ rule to determine whether 95
minutes is an outlier.
\item Explain why the mean AI-tutor minutes would be larger than the
median for Table A.
\end{enumerate}

\subsection*{Unit 2: Exploring Two-Variable Data}
\begin{enumerate}[resume,leftmargin=*]
\item Use Table D to compute the conditional relative frequency of
regular AI use within each school culture.
\item Does Table D suggest an association between school culture and
actual AI use? Explain using conditional relative frequencies.
\item Interpret the slope $0.40$ in Table E in context.
\item Interpret $r = 0.62$ in context.
\item Find and interpret $r^2$.
\item Predict the gain score for a student who uses the AI tutor for
40 minutes per day.
\item Explain why it would be risky to use the regression line to
predict the gain score for a student who uses AI for 180 minutes
per day.
\end{enumerate}

\subsection*{Unit 3: Collecting Data}
\begin{enumerate}[resume,leftmargin=*]
\item Identify the experimental units, treatments, and response
variables in the six-week study.
\item Explain why random assignment within each school improves the
study.
\item Explain why blocking by school would be sensible in this study.
\item Can the randomized comparison in Table B support a
cause-and-effect conclusion about AI access? Explain.
\item Can the comparison in Table C support a cause-and-effect
conclusion about actual AI use? Explain carefully.
\item Identify one possible source of bias if students self-report
their AI minutes.
\item Explain how noncompliance can contaminate the interpretation of
a treatment effect.
\end{enumerate}

\subsection*{Unit 4: Probability, Random Variables, and Probability Distributions}
\begin{enumerate}[resume,leftmargin=*]
\item Using Table D, find $P(\text{regular AI use})$.
\item Using Table D, find $P(\text{open AI-tutor culture})$.
\item Using Table D, find $P(\text{regular AI use} \mid \text{open AI-tutor culture})$.
\item Are regular AI use and open AI-tutor culture independent in
Table D? Justify numerically.
\item Suppose 5 students are randomly selected from a population
where $p = 0.32$ for regular AI-tutor use. Find the probability
exactly 2 are regular AI-tutor users.
\item Let $X$ be the number of regular AI-tutor users among 5
randomly selected students with $p = 0.32$. Find $E(X)$ and
$\text{SD}(X)$.
\item Find the probability that the first regular AI-tutor user
appears on the fourth student selected.
\end{enumerate}

\subsection*{Unit 5: Sampling Distributions}
\begin{enumerate}[resume,leftmargin=*]
\item For samples of size 120 from a population with true pass
proportion $p = 0.50$, describe the approximate sampling
distribution of $\hat{p}$. Check conditions.
\item If the true mean gain under the standard-resource condition is
10 points with population SD 12, describe the approximate sampling
distribution of the sample mean gain for $n = 120$.
\item Under the null hypothesis that the two pass proportions are
equal, explain why the observed difference in pass rates from
Table B might or might not be surprising.
\item Explain why larger sample sizes make it easier to detect
smaller real differences.
\end{enumerate}

\subsection*{Unit 6: Inference for Categorical Data: Proportions}
\begin{enumerate}[resume,leftmargin=*]
\item Construct a 95\% confidence interval for the proportion of
AI-access students who passed the AP-style practice exam.
\item Construct a 95\% confidence interval for the difference in pass
proportions, $p_{AI} - p_{\text{standard}}$, using Table B.
\item Interpret the interval from Question 32 in context.
\item Test whether the AI-access condition produced a higher pass
proportion than the standard-resource condition.
\item Explain the meaning of the p-value from Question 34 in context.
\item Describe a Type I error and a Type II error for the test in
Question 34.
\end{enumerate}

\subsection*{Unit 7: Inference for Quantitative Data: Means}
\begin{enumerate}[resume,leftmargin=*]
\item Construct a 95\% confidence interval for the mean gain in the
AI-access condition.
\item Construct a 95\% confidence interval for the difference in mean
gains, $\mu_{AI} - \mu_{\text{standard}}$, using Table B.
\item Interpret the interval from Question 38 in context.
\item Test whether the AI-access condition produced a greater mean
gain than the standard-resource condition.
\item Explain why the randomized design matters when interpreting the
difference in mean gains.
\item Suppose a critic says, ``Students in the AI-access group were
already stronger.'' Use Table B to respond statistically.
\end{enumerate}

\subsection*{Unit 8: Inference for Categorical Data: Chi-Square}
\begin{enumerate}[resume,leftmargin=*]
\item Use Table D to state appropriate hypotheses for a chi-square test.
\item Calculate the expected count for restrictive culture and
regular AI use.
\item Verify that the conditions for a chi-square test are met.
\item The chi-square test statistic for Table D is approximately
$43.62$. Find the degrees of freedom and approximate p-value
category.
\item State a conclusion in context.
\item Identify which cells contribute most to the chi-square
statistic and explain what that means in context.
\end{enumerate}

\subsection*{Unit 9: Inference for Quantitative Data: Slopes}
\begin{enumerate}[resume,leftmargin=*]
\item State hypotheses for testing whether average daily AI-tutor
minutes are linearly associated with gain score.
\item Use Table E to calculate the test statistic for the slope.
\item Find the degrees of freedom.
\item State a conclusion in context.
\item Construct a 95\% confidence interval for the slope.
\item Interpret the interval in context.
\item Explain why regression evidence alone is weaker causal evidence
than the randomized comparison in Table B.
\end{enumerate}

\subsection*{Final Synthesis}
\begin{enumerate}[resume,leftmargin=*]
\item A school administrator says, ``The AP-style scores prove
AI-supported instruction works.'' Explain why that statement is
too strong.
\item A teacher says, ``The AP-style scores prove AI-supported
instruction fails whenever students do not use it correctly.''
Explain why that statement is also too strong.
\item Distinguish between the \textbf{intent-to-treat effect} of AI
access and the \textbf{on-treatment association} for students who
actually used AI regularly.
\item Write a statistically careful conclusion about what the full
dataset does and does not show.
\item Propose a better follow-up study that could distinguish
AI-tutor effects from student motivation, compliance, teacher
culture, and phone-policy effects.
\end{enumerate}

\newpage
\section{Teacher Notes / Answer Key}

\begin{keybox}
Worked answers below. Approximations are AP-appropriate; minor
rounding differences are acceptable.
\end{keybox}

\begin{enumerate}[leftmargin=*]
\item Condition, school culture, and passed exam are categorical.
AI minutes and gain score are quantitative.

\item A dotplot, histogram, or boxplot is appropriate. The
distribution is right-skewed with high values at 75 and 95
minutes. A typical value is around 25 to 30 minutes. The data
range from 0 to 95 minutes.

\item Median $27.5$; $Q_1 = 11$; $Q_3 = 47.5$; IQR $= 36.5$; range $= 95$.

\item Lower fence $11 - 1.5(36.5) = -43.75$; upper fence
$47.5 + 1.5(36.5) = 102.25$. The value 95 is \emph{not} an outlier
by the $1.5 \times \text{IQR}$ rule.

\item The large high-end values create a right tail, and the mean is
pulled toward that tail.

\item Restrictive regular-use rate: $21/120 = 0.175$. Open-culture
regular-use rate: $66/120 = 0.55$.

\item Yes. Regular use is much higher in open AI-tutor culture, while
no use is much higher in restrictive culture: $60/120 = 0.50$
versus $20/120 = 0.167$.

\item For each additional AI-tutor minute per day, predicted gain
increases by $0.40$ points, on average.

\item There is a moderately strong positive linear association
between AI-tutor minutes and gain score.

\item $r^2 = 0.62^2 = 0.3844$. About 38.4\% of variation in gain
score is explained by the linear relationship with AI-tutor
minutes.

\item $5.8 + 0.40(40) = 21.8$ points.

\item 180 minutes is likely outside the observed range, so this is
extrapolation. The linear pattern may not continue that far.

\item Experimental units are AP Statistics students. Treatments are
AI access and standard resources. Response variables include
posttest score, gain score, pass status, and engagement.

\item Random assignment reduces systematic differences between
groups, making the groups more comparable except for the assigned
condition.

\item Blocking by school controls for school-level differences such
as teacher culture, phone policy, scheduling, prior preparation,
and local norms.

\item Yes, the randomized comparison can support a cause-and-effect
conclusion about \emph{being assigned} AI access, assuming the
study was carried out properly.

\item No, not by itself. Actual use was not randomly assigned; it may
be confounded with motivation, prior skill, family support,
teacher encouragement, or school culture.

\item Students may overreport use to look diligent, underreport use
if they fear punishment, or misremember their actual minutes.

\item Noncompliance means the assigned treatment and the received
treatment differ. The assigned-group estimate can understate,
distort, or complicate the effect of actually using the tool.

\item $87/240 = 0.3625$.

\item $120/240 = 0.5$.

\item $66/120 = 0.55$.

\item Not independent because $P(\text{regular}) = 0.3625$, but
$P(\text{regular} \mid \text{open}) = 0.55$.

\item $\binom{5}{2}(0.32)^2(0.68)^3 \approx 0.322$.

\item $E(X) = np = 5(0.32) = 1.6$;
$\text{SD}(X) = \sqrt{np(1-p)} = \sqrt{5(0.32)(0.68)} \approx 1.04$.

\item $(0.68)^3(0.32) \approx 0.101$.

\item Mean $0.50$; SD $\sqrt{0.5 \cdot 0.5 / 120} \approx 0.0456$;
approximately normal because expected successes and failures are
both at least 10.

\item Mean $10$; SD $12/\sqrt{120} \approx 1.095$; approximately
normal by the CLT because $n = 120$ is large.

\item The observed difference is $69/120 - 47/120 = 0.183$. Whether
this is surprising depends on the null sampling distribution.
Later inference shows it is unlikely under equal pass proportions.

\item Larger sample sizes reduce standard error, so the sampling
distribution is narrower. A real difference has a better chance
of producing a large test statistic.

\item $\hat{p} = 69/120 = 0.575$;
$SE = \sqrt{0.575 \cdot 0.425 / 120} \approx 0.0451$;
95\% CI $\approx 0.575 \pm 1.96(0.0451) = (0.486,\ 0.664)$.

\item Difference $= 0.575 - 0.392 = 0.183$;
$SE = \sqrt{\frac{0.575 \cdot 0.425}{120} + \frac{0.392 \cdot 0.608}{120}} \approx 0.0635$;
95\% CI $\approx (0.059,\ 0.308)$.

\item We are 95\% confident the AI-access pass proportion is about
5.9 to 30.8 percentage points higher than the standard-resource
pass proportion.

\item $H_0\!:\ p_{AI} = p_{\text{standard}}$;
$H_a\!:\ p_{AI} > p_{\text{standard}}$.
Pooled proportion $= 116/240 = 0.483$.
$SE = \sqrt{0.483 \cdot 0.517 \cdot (\tfrac{1}{120} + \tfrac{1}{120})} \approx 0.0645$.
$z \approx 2.84$; p-value $\approx 0.0023$; reject $H_0$.

\item If true pass proportions were equal, the probability of getting
a difference at least this large in favor of AI access by random
chance is about 0.0023.

\item Type I: concluding AI access improves pass rate when it does
not. Type II: failing to detect a real improvement from AI access.

\item $18.6 \pm 1.98(12.4/\sqrt{120}) \approx (16.36,\ 20.84)$.

\item Difference $= 8.8$;
$SE = \sqrt{12.4^2/120 + 11.7^2/120} \approx 1.56$;
95\% CI $\approx (5.72,\ 11.88)$.

\item We are 95\% confident AI access increases mean gain by about
5.7 to 11.9 points compared with standard resources.

\item $H_0\!:\ \mu_{AI} = \mu_{\text{standard}}$;
$H_a\!:\ \mu_{AI} > \mu_{\text{standard}}$.
$t \approx 8.8/1.56 = 5.65$; p-value is very small; reject $H_0$.

\item Randomized design makes it more reasonable to attribute the
difference in mean gains to assigned AI access rather than
preexisting group differences.

\item The pretest means are similar: $48.2$ versus $48.9$, so the
data do not support the claim that the AI-access group started
stronger on average.

\item $H_0$: school culture and AI-use category are independent.
$H_a$: school culture and AI-use category are associated.

\item Expected count for restrictive culture and regular AI use:
$120 \cdot 87 / 240 = 43.5$.

\item Expected counts are $43.5$, $36.5$, and $40$ in each row, all
at least 5. The data must also come from an appropriate random
or well-designed study.

\item Degrees of freedom $(2-1)(3-1) = 2$. With $\chi^2 \approx 43.62$,
p-value is less than $0.001$.

\item Reject independence. There is strong evidence of an association
between school culture and AI-use category.

\item The regular-use and no-use cells contribute most. Restrictive
culture has far fewer regular users and far more nonusers than
expected; open culture has far more regular users and far fewer
nonusers than expected.

\item $H_0\!:\ \beta = 0$; $H_a\!:\ \beta > 0$ if testing for
positive association, or $H_a\!:\ \beta \neq 0$ if testing for any
linear association.

\item $t = 0.40 / 0.061 \approx 6.56$.

\item $df = n - 2 = 78$.

\item Reject $H_0$; there is strong evidence of a positive linear
association between AI-tutor minutes and gain score.

\item Using approximately $t^* = 1.99$:
CI $= 0.40 \pm 1.99(0.061) \approx (0.279,\ 0.521)$.

\item Each additional AI-tutor minute per day is associated with an
average gain increase of about $0.279$ to $0.521$ points.

\item Regression here uses actual minutes, which were not randomly
assigned. Motivation and other variables could confound the
relationship.

\item It is too strong because the study directly supports a causal
claim about \emph{assigned access}, not a universal claim that
AI-supported instruction always works. Implementation, student
use, and assessment conditions still matter.

\item It is too strong because misuse or nonuse of a tool does not
prove the tool is ineffective for students who use it as intended.
Noncompliance contaminates the outcome.

\item The intent-to-treat effect compares assigned AI access with
assigned standard resources. The on-treatment association
compares students by what they actually did, such as regular AI
use, but that comparison is more vulnerable to self-selection.

\item A careful conclusion: in this synthetic study, randomized AI
access is associated with higher mean gains and pass rates, and
actual AI use is positively associated with gain. However, actual
use is confounded with student motivation and school culture, so
the cleanest causal conclusion is about access under the study
design.

\item A stronger follow-up study would randomize schools or classes
to clearly defined phone/AI policies, track actual AI use with
logs, require common in-class no-AI assessments, block by school
and teacher, measure prior achievement and motivation, and
compare both intent-to-treat and per-protocol results.
\end{enumerate}

\subsection*{Final Synthesis Guidance}
A strong final answer should say:
\begin{itemize}[leftmargin=*]
  \item The randomized evidence supports a causal claim about
  \textbf{AI access}, not necessarily pure actual AI use.
  \item The actual-use evidence is stronger as an association than as
  a causal claim because of self-selection.
  \item School culture is associated with usage, so local anti-AI or
  pro-AI norms can contaminate interpretation.
  \item AP-style outcomes alone are insufficient if student effort,
  buy-in, or noncompliance differs across groups.
  \item The cleanest conclusion is that AI access, paired with
  independent demonstration requirements, appears to improve
  performance in this synthetic study, but further randomized work
  should separate access, actual use, student motivation, teacher
  implementation, and school culture.
\end{itemize}

\newpage
% ============================================================
\part{Investigative Task: Evaluating an AI-Tutor Pedagogy}
% ============================================================

\begin{scenario}
A high school AP Statistics teacher in Massachusetts implemented an
AI-tutor-supplemented curriculum during the 2024--25 school year. The
class had $n = 22$ students. At year's end, \textbf{all 22 scored a 1}
on the AP exam. The national 1-rate that year was
$\mathbf{23.7\%}$ (College Board, 2025 score distribution; the 5-rate
that year was $17.0\%$).
The teacher reports that 18 of the 22 students ``boycotted'' the exam
(left it largely blank as protest); the remaining 4 attempted it.

Separately, a randomized controlled trial published in
\emph{Scientific Reports} (Kestin, Miller, Klales, Milbourne, \&
Ponti, 2025; \href{https://www.nature.com/articles/s41598-025-97652-6}{DOI 10.1038/s41598-025-97652-6};
$N = 194$ Harvard undergraduates in introductory physics, randomized
\textbf{crossover} design over two $\sim$75-minute lessons) compared
an AI tutor against a research-based \textbf{active-learning}
classroom (peer instruction, small groups, instructor feedback) on
identical content.

Reported result: AI-tutor post-score \textbf{median} $= 4.5$
($n = 142$), active-learning post-score \textbf{median} $= 3.5$
($n = 174$). Inferential test: \textbf{Mann--Whitney rank-sum}
$z = -5.6$, $p < 10^{-8}$. Regression-based effect size
$d \approx 0.63$ (biased low by ceiling effects); ceiling-adjusted
quantile estimates range $0.73$ to $1.3$ standard deviations. The AI
arm also reported higher engagement and used less median time
(49 vs.\ 60 minutes).

The teacher's father, citing the all-1s outcome, argues
\emph{``the data shows your method failed.''}
\end{scenario}

\section{(a) Sampling and external validity \textnormal{\small(Unit 3)}}

The teacher's class is \textbf{not} a random sample from any larger
population. Identify (i) the population the local result can validly
describe and (ii) the population the RCT can validly describe.
Explain why averaging the two as if equally weighted commits a
\textbf{selection bias} error.

\begin{expectedbox}
Local sample $=$ volunteer convenience sample, single classroom,
single teacher in a single Massachusetts district, $n = 22$ ---
describes only itself. RCT $=$ randomized within a defined frame
(Harvard undergrads in an introductory physics course) --- supports
causal inference within that frame and provides reasonable
generalization to similar learners. The RCT does \emph{not}
automatically generalize to high-school AP Statistics, but its
internal validity is strong. The local sample is non-random and
confounded with political/cultural variables; it cannot be pooled as
evidence against the RCT.
\end{expectedbox}

\section{(b) Inference on the local outcome \textnormal{\small(Units 4 \& 6)}}

Test $H_0: p = 0.237$ vs. $H_a: p > 0.237$ where $p$ is the
probability a student in this class scores a 1.

\textbf{Step 1: check conditions for the one-proportion z-test.}
Large Counts requires $np_0 \geq 10$ and $n(1 - p_0) \geq 10$.
\[
np_0 = 22(0.237) = 5.21 < 10
\]
\textbf{The Large Counts condition fails.} The sampling distribution
of $\hat p$ is not approximately normal under $H_0$, so a
one-proportion z-test is \emph{not appropriate} here.

\textbf{Step 2: use the exact binomial test.}
Under $H_0$, the count of 1s satisfies $X \sim \text{Binomial}(22,\ 0.237)$.
The observed value is $X = 22$.
\[
\text{p-value} = P(X = 22 \mid X \sim \text{Bin}(22,\ 0.237))
= (0.237)^{22} \approx 1.76 \times 10^{-14}
\]

\textbf{Reject $H_0$.} The 1-rate in this class is astronomically
higher than the national rate.

\begin{notebox}
\textbf{Calculator / software:}
\begin{itemize}[leftmargin=*,nosep]
  \item TI-84: \code{1 - binomcdf(22, 0.237, 21)}
  \item Excel: \code{=1 - BINOM.DIST(21, 22, 0.237, TRUE)}
  \item R: \code{binom.test(22, 22, p = 0.237, alternative = "greater")}
\end{itemize}
\end{notebox}

The rejection above tells us the class's outcome is wildly
inconsistent with national averages. It does \emph{not}, by itself,
support the father's causal claim. Why not? See part (c).

\section{(c) Intent-to-treat vs. per-protocol; identifying the estimand \textnormal{\small(Units 3 \& 6)}}

Of 22 students, 18 deliberately did not attempt the exam. Define:
\begin{itemize}[leftmargin=*]
  \item \textbf{Intent-to-treat (ITT) estimand:} mean exam score for
  \emph{all} assigned students. Includes non-compliers.
  \item \textbf{Per-protocol (PP) estimand:} mean exam score for
  students who \emph{actually engaged} with both the instruction and
  the assessment.
\end{itemize}
The all-1s figure is an ITT measure with $\sim$82\% non-compliance.
Argue that the ITT estimate here measures \textbf{student compliance
with the assessment}, not the causal effect of AI-tutor instruction
on statistical reasoning skill.

\begin{notebox}
\textbf{Analogy:} a drug RCT in which 82\% of the treatment arm
refuses to swallow the pill cannot be used to claim the drug is
ineffective.
\end{notebox}

\textbf{The relevant on-treatment data} are: in-class Blooket
performance, FRQ-style work submitted under observation, verbal
defenses of reasoning. The father's argument confuses ITT
non-compliance with causal failure.

\section{(d) Inference on the published RCT \textnormal{\small(Unit 7 interpretation)}}

The Kestin et al. (2025) study did \textbf{not} use a two-sample
t-test. The published analysis is:

\begin{itemize}[leftmargin=*]
  \item \textbf{Design:} randomized counterbalanced \emph{crossover}
  --- every eligible student experienced both conditions across two
  $\sim$75-minute lessons (surface tension; fluid flow). Randomization
  occurred at the peer-instruction-group level (clusters of 2--3),
  not strictly at the individual level.
  \item \textbf{Comparator:} a research-based \emph{active-learning}
  classroom (peer instruction, small groups, instructor feedback),
  \emph{not} traditional lecture. This makes the result more striking
  to specialists.
  \item \textbf{Statistics:} medians 4.5 (AI-tutor, $n = 142$) vs.\
  3.5 (active learning, $n = 174$). Mann--Whitney rank-sum $z = -5.6$,
  $p < 10^{-8}$.
  \item \textbf{Effect size:} regression-based $d \approx 0.63$,
  acknowledged biased low by post-test ceiling effects;
  ceiling-adjusted quantile estimates range $0.73$ to $1.3$
  standard deviations.
  \item \textbf{Auxiliary findings:} higher self-reported engagement
  in the AI arm; median time-on-task 49 min (AI) vs.\ 60 min
  (active learning).
\end{itemize}

\textbf{Conclusion:} The AI-tutor condition produced significantly
higher median post-test performance than a high-quality active-learning
condition, with a large effect size, in less time. Causal direction
is licensed within the study's frame because randomization was
preserved across the crossover. Generalization to high-school AP
Statistics requires care: the content was first-encounter physics at
Bloom's lower-to-middle levels, the AI used pre-written step-by-step
solutions embedded in prompts, and no independent replication has
been published as of this writing.

\begin{notebox}
\textbf{What students should be able to interpret from these
statistics:} a Mann--Whitney test compares ranks (not means), is
robust to non-normal distributions, and is appropriate when the
outcome is ordinal or skewed by ceiling/floor effects. The huge
$|z| = 5.6$ and effect size in the $0.63$--$1.3$ range together
indicate a result that is both \emph{statistically} and
\emph{practically} significant.
\end{notebox}

\section{(e) Power and the ``we don't have the data yet'' claim \textnormal{\small(beyond AP --- enrichment)}}

\begin{notebox}
\textbf{Beyond the AP CED.} Formal power calculation is enrichment;
include it for context, not for exam preparation.
\end{notebox}

For a two-sample comparison with $d \approx 1.0$ at $\alpha = 0.05$
and $n_1 = n_2 = 97$, approximate power is:
\[
\text{power} \approx P\!\left(Z > 1.96 - 1.0\sqrt{97/2}\right)
= P(Z > 1.96 - 6.96) \approx 1.000
\]
Even using the more conservative $d = 0.63$, power exceeds $0.99$
at this sample size. The claim ``we don't have the data'' is
equivalent to either (i) being unaware of high-powered evidence that
already exists, or (ii) treating $\alpha$ thresholds asymmetrically
(require evidence for AI, accept tradition by default). Both are
inferential errors, not neutral positions.

\section{(f) Confounding in the local sample \textnormal{\small(Unit 3)}}

List confounders in the local AP-score outcome that are \textbf{not}
controlled in the local ``study'':
\begin{itemize}[leftmargin=*]
  \item Self-selection into the class
  \item Single-teacher / single-school effect (no replication)
  \item Local cultural norms about AI and assessment compliance
  \item Knowledge that the score itself was being used as social currency
  \item Teacher reputation effect (admin/peer signaling about the teacher)
  \item Student political identification influencing assessment effort
\end{itemize}
The Kestin RCT controls these by design via randomization within a
single course. The local outcome does not. The local outcome is
therefore not a valid test of the treatment effect even before the
non-compliance issue.

\section{(g) Bastani et al.\ (2025) and the ``crutch effect'' \textnormal{\small(Unit 3 + Unit 7)}}

\begin{cautionbox}
\textbf{The most relevant K--12 finding to date.}
Bastani, Bastani, Sungu, Ge, Kabakc{\i}, \& Mariman (2025),
``Generative AI without guardrails can harm learning,''
\emph{PNAS} 122(26), e2422633122.

\textbf{Setting:} $\sim$1{,}000 high-school math students at TED Ankara
Koleji, grades 9--11, three randomized arms:
(1) GPT Base (unrestricted GPT-4),
(2) GPT Tutor (Socratic guardrails designed by teachers),
(3) control (no AI).

\textbf{During practice with AI access:} both AI arms outperformed
control by 48\% (Base) and 127\% (Tutor).

\textbf{On a later \emph{unassisted} exam (AI removed):} the GPT Base
arm scored \textbf{17\% lower than control} --- a statistically
significant deficit. The scaffolded GPT Tutor arm eliminated the harm
but produced no significant gain over control. Students did not
perceive the harm; they thought they had learned.
\end{cautionbox}

\textbf{Interpretation.} Bastani is the strongest cautionary K--12
evidence available. It is not a refutation of AI-as-tutor; it is a
refutation of \emph{unguarded answer-engine use during practice}.
The harm condition was: students could ask GPT-4 anything and receive
direct answers, no scaffolding, no required independent demonstration.

\textbf{Why this strengthens the refined thesis rather than weakening
it:} the classroom design described in this document --- AI permitted
during learning, mandatory handwritten reasoning, verbal defense, and
no-AI in-class assessments --- is the \emph{exact inverse} of the
Bastani harm condition. Bastani is therefore evidence for the
scaffolded-tutoring model, not against it. A defensible curriculum
must engage Bastani directly:

\begin{itemize}[leftmargin=*]
  \item Unscaffolded AI access $\to$ short-run gains, transfer
  deficit on no-AI assessment.
  \item Scaffolded AI tutoring + required independent demonstration
  $\to$ gains preserved on no-AI assessment.
  \item Prohibition $\to$ no gains, no transfer effect, but also no
  upside.
\end{itemize}

The refined thesis: \emph{the design of AI use determines whether it
becomes high-quality tutoring or a crutch.}

\section{(h) Chi-square test for differential response \textnormal{\small(Unit 8)}}

Suppose the teacher records the following $2 \times 2$ table from the
\emph{next} cohort ($n = 30$), classifying students by prior
engagement and post-tutor performance:

\begin{center}
\small
\begin{tabular}{lrrr}
\toprule
 & Improved $\geq$ 1 letter grade & Did not improve & Total \\
\midrule
Previously disengaged & 9 & 3  & 12 \\
Previously engaged    & 4 & 14 & 18 \\
\midrule
\textbf{Total}        & \textbf{13} & \textbf{17} & \textbf{30} \\
\bottomrule
\end{tabular}
\end{center}

Test $H_0$: improvement is independent of prior-engagement status.

Expected counts: e.g., $E_{11} = 12 \cdot 13 / 30 = 5.2$.
Computed $\chi^2 \approx 8.6$, $df = 1$, $p \approx 0.003$.

\textbf{Reject independence.} AI-tutor benefit in this cohort is
\textbf{not uniform}: marginal gain is concentrated in students the
prior system underserved. This pattern is consistent with the broader
tutoring literature, in which the largest gains accrue to students
furthest from the group-paced mean.

\begin{notebox}
\textbf{Note on the tutoring literature.} Bloom's (1984) ``2-sigma''
claim is widely cited but rests on two unpublished doctoral
dissertations and has not been reproduced in subsequent
meta-analyses. Cohen, Kulik, \& Kulik (1982) reported a pooled
tutoring effect of $d = 0.33$; VanLehn (2011) found human and
intelligent-tutor effects both near $d \approx 0.76$; Nickow,
Oreopoulos, \& Quan (2024) reported pooled effect $d = 0.288$.
The tutoring effect is real and robust, but at moderate magnitudes,
not 2 standard deviations. Treat Bloom as historical aspiration, not
as predictive evidence.
\end{notebox}

\section{(i) Linear regression with interaction \textnormal{\small(beyond AP --- enrichment)}}

\begin{notebox}
\textbf{Beyond the AP CED.} The AP Statistics framework covers
single-predictor linear regression and inference for slope. Multiple
regression with interaction terms is enrichment. Include for
intuition; do not assess.
\end{notebox}

Conceptually, fit:
\[
\widehat{\text{Score}} = \beta_0 + \beta_1(\text{AI hours})
  + \beta_2(\text{prior engagement})
  + \beta_3(\text{AI hours} \times \text{prior engagement})
\]
A finding where $\hat\beta_1 > 0$ and $\hat\beta_3 < 0$ would mean:
AI hours raise score on average, but the \emph{marginal} benefit
shrinks for already-engaged students. This formalizes the qualitative
claim: \textbf{AI as tutor disproportionately helps students with the
most slack between actual and potential performance}, rather than
uniformly boosting everyone.

\section{(j) Synthesis: stating the defensible conclusion \textnormal{\small(Skills 1.E, 4.B)}}

Write a paragraph, in AP-rubric style, that:
\begin{enumerate}[leftmargin=*]
  \item Identifies the \textbf{causal question} (``Does scaffolded
  AI-tutor instruction increase statistical reasoning skill relative
  to either prohibition or unscaffolded AI access?'').
  \item Identifies \textbf{which evidence is admissible} for that
  question (Kestin RCT and Bastani RCT --- yes, with caveats; local
  ITT outcome under boycott --- no).
  \item States the \textbf{direction and magnitude} of the
  admissible evidence (Kestin: $z = -5.6$, $p < 10^{-8}$, $d$ in the
  $0.63$--$1.3$ range; Bastani: $-17\%$ deficit on no-AI exam under
  unscaffolded use).
  \item Identifies the \textbf{failure mode} of the local
  ``evidence'' (selection bias, confounding, near-total
  non-compliance).
  \item Concludes that the empirical claim ---
  \emph{``Scaffolded AI tutoring with required independent
  demonstration produces faster skill acquisition than either
  prohibition or unscaffolded answer-engine use, with the largest
  marginal gains for previously underserved students''} --- is
  \textbf{supported by the data available as of 2026}, and that
  arguments to the contrary based on the local outcome rely on
  inferential errors a Unit 3 / Unit 6 student is expected to
  identify.
\end{enumerate}

\section*{Skills covered}

\begin{center}
\small
\begin{tabular}{cl}
\toprule
Part & AP Stats skill \\
\midrule
(a) & Sampling, external validity, generalizability \\
(b) & Large Counts condition, exact binomial test \\
(c) & Estimand definition, ITT vs PP, interpretation under non-compliance \\
(d) & Interpreting published nonparametric inference (Mann--Whitney), effect size \\
(e) & Power, Type II error, asymmetric evidentiary standards \emph{(beyond AP)} \\
(f) & Confounding, randomization as causal license \\
(g) & Engaging cautionary RCT evidence (Bastani) \\
(h) & Chi-square independence, differential treatment effect \\
(i) & Regression interaction, marginal effect \emph{(beyond AP)} \\
(j) & Scope of inference, synthesis, communication \\
\bottomrule
\end{tabular}
\end{center}

\vspace{1em}
\begin{notebox}
\textbf{Pedagogical point:} a student who can correctly answer
(a)--(j) can reconstruct the argument from first principles, without
needing to share the author's priors. The defense is not ``trust me''
--- it is ``here is the test, and here is what the rubric says.''
\end{notebox}

\section*{References}

\begin{itemize}[leftmargin=*]
\item Bastani, H., Bastani, O., Sungu, A., Ge, H., Kabakc{\i}, \"O.,
\& Mariman, R. (2025). Generative AI without guardrails can harm
learning: Evidence from high school mathematics. \emph{PNAS},
122(26), e2422633122.
\item Bloom, B. S. (1984). The 2 sigma problem. \emph{Educational
Researcher}, 13(6), 4--16.
\item Cohen, P. A., Kulik, J. A., \& Kulik, C. C. (1982). Educational
outcomes of tutoring: A meta-analysis of findings. \emph{American
Educational Research Journal}, 19(2), 237--248.
\item College Board. (2025). \emph{AP Statistics score distributions}.
Retrieved from AP Central.
\item Kestin, G., Miller, K., Klales, A., Milbourne, T., \& Ponti, G.
(2025). AI tutoring outperforms in-class active learning: An RCT
introducing a novel research-based design in an authentic
educational setting. \emph{Scientific Reports}, 15, 17458.
\href{https://doi.org/10.1038/s41598-025-97652-6}{DOI 10.1038/s41598-025-97652-6}.
\item Nickow, A., Oreopoulos, P., \& Quan, V. (2024). The promise of
tutoring for PreK--12 learning: A systematic review and
meta-analysis of the experimental evidence. \emph{American
Educational Research Journal}, 61(1), 74--107.
\item VanLehn, K. (2011). The relative effectiveness of human
tutoring, intelligent tutoring systems, and other tutoring systems.
\emph{Educational Psychologist}, 46(4), 197--221.
\end{itemize}

\end{document}
